---
title: "6. Simulate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Simulate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=TRUE, results='hide', message=FALSE}
n_clocks=3
n_events=8
purity=0.9 
coverage=100 
epsilon=0.20 
seed = 123 
tolerance = 0.001
max_attempts = 2
INIT = TRUE
min_mutations_number = 3


cat("coverage: ", coverage)

mu = 1e-4 # mutation rate
cat("mutation rate: ", mu)

w = 1e-2 # cell division rate
cat("cell division rate: ", w)

l = 2e7 # length of the segment
cat("length of the segment: ", l)

time_interval = 7
cat("time_interval: ", time_interval)


res_simulate <- tickTack::get_simulation_tickTack(number_clocks=n_clocks, 
                                               number_events=n_events, 
                                               purity=purity, 
                                               coverage=coverage, 
                                               epsilon=epsilon, 
                                               seed = seed)
  
  data_simulation = as.data.frame(res_simulate$data_simulation) %>% mutate(chr=   1:length(res_simulate$data_simulation$taus))
  x = res_simulate$x
  
  df = x$cna %>% left_join(data_simulation)
  
 # timing inference ticktack hierarchical

  x <- tickTack::fit_h(x, max_attempts=max_attempts, INIT=INIT, tolerance = tolerance)
  
  
  results_simulated <- x$results_timing
  results_model_selection <- tickTack::model_selection_h(results_simulated)
  best_K <- results_model_selection$best_K



```


### Results

The `results` object that is returned together with the CNAqc input object contains four components: `data`, `draws_and_summary`, `log_lik_matrix_list` and `elbo_iterations`. 

```{r echo=TRUE}
# View summary for a specific K, here K = 2
results <- x$results
```

### Interpreting the output

We can inspect the main output of interest to understand the timing of clonal peaks.
`results$draws_and_summary` contains:
- **`draws`** the draws from the approximate posterior distribution of the taus and weights;
- **`summary`** a summary with the main statistics of the approximate posterior distributions;
- **`summarized_results`** represents the clock assignment, a tibble with the estimate of taus for each segment with a copy number event that has been included in the hierarchical inference

```{r echo=TRUE}
# View summary for a specific K, here K = 2
results$draws_and_summary[[2]]$summary

# View detailed summarized results for a specific K, here K = 2
results$draws_and_summary[[2]]$summarized_results
```



### Obtain the best K with model_selection_h
W e can run the `model_selection_h` function to obtain the scores for each inference performed with a different K and take the one with best ICL score if the BIC score prefer 2 components instead of 1, otherwise choose 1 as best K. The function takes as input the `results` and `n_components` and outputs the `best_K` and the corresponding `best_fit` together with the `model_selection_tibble` and the `entropy_list` used to evaluate  the ICL score.

```{r echo=TRUE, results='hide', message=FALSE}
results_model_selection <- tickTack::model_selection_h(results, n_components = 0)

best_K <- results_model_selection$best_K
model_selection_tibble <- results_model_selection$model_selection_tibble
entropy <- results_model_selection$entropy_list

```


## Visulizing the output

The results can be viewed is genome-wise perspective using the `tickTack::plot_timing_h` function.

```{r, fig.align='center', fig.width=6, fig.height=4}
tickTack::plot_timing_h(results, 2)
```



## Visualize distributions of draws from the approximate posterior 
The approximate posterior distributions can be viewed using the `tickTack::plot_posterior_clocks_h`  and `tickTack::plot_posterior_weights_h`  functions, that internally use functions from Bayesplot.


```{r, fig.align='center', fig.width=6, fig.height=4}
posterior_clocks <- tickTack::plot_posterior_clocks_h(results, 2)
posterior_weights <- tickTack::plot_posterior_weights_h(results, 2)

```


## Visualize  the behavior of the ELBO during the inference


```{r, fig.align='center', fig.width=6, fig.height=5}
K = nrow(results_model_selection$model_selection_tibble)

p_elbo <- list()
for (i in 1:K){
  p_elbo[[i]] <- tickTack::plot_elbo_h(results$elbo_iterations[[i]]) + ggplot2::ggtitle(paste0("K = ", i))
}
p_elbo <- gridExtra::grid.arrange(grobs = p_elbo, ncol = 2)  #add global title
p_elbo

```


<!-- ## Visualize model selection scores -->

<!-- ```{r} -->
<!-- #  -->
<!-- # p_model_selection <- tickTack::plot_model_selection_h(model_selection_tibble, best_K) -->
<!-- # p_model_selection -->
<!-- #  -->
<!-- # cat("Best K =",best_K) -->
<!-- ``` -->


## Visualize all the inference results for each K

```{r, fig.align='center', fig.width=6, fig.height=5, message=FALSE}
plot_model_selection_inference <- list()
for (i in 1:K){
  plot_model_selection_inference[[i]] <- tickTack::plot_timing_h(results, i) + ggplot2::ggtitle(paste0("K = ", i))
}
plot_model_selection_inference <- gridExtra::grid.arrange(grobs = plot_model_selection_inference, ncol = 2) #add global title
plot_model_selection_inference

```
